### 2010-03-17 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: uniformize spacing in arrays
Desc@ICFP: composed by -> composed of
Desc@ICFP: uniformize EnumT definitions
Desc@ICFP: uniformize sum-of-product and sigmas-of-sigmas

In the simply-typed world, the essence of data-type definition can be
summed up by the expression ``sum-of-product''. A data-type is defined
by a finite sum -- a choice -- of constructors, each of them is
composed by a product -- a tuple -- of arguments. This form governs
the grammar of data-type definition in languages such as Haskell or
OCaml.

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of ``sum-of-product''
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a see of $\Sigma$s, our data-types have little
chance of survival. \note{Ridiculous tentative of poetry. I was
  really tired when I wrote that.}

@@

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& X}        &  \Unit                                       \\
\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
\end{array}
}
\end{array}
\]

@@

\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
}\]

@@

\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: X \mapsto \DSigma{(\EnumT [ \ListNil, \ListCons ])}
                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]

@@

\[\stk{
\TreeD : \Set \To \Desc \\
\begin{array}{@{}ll}
\TreeD \: X \mapsto \DSigma{ & (\EnumT [ \TreeLeaf, \TreeNode ]) \\}
                           { & [ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
\end{array}
}\]

@@

Again, this is a clear reminiscence of the ``sum-of-products'' style:
we have a finite sum of constructors. Of course, every description can
be expressed in this style, using a singleton as single constructor if
necessary. A tagged description can trivially be turned into a
full-blown description. We note $\toDesc{TD}$ the description computed
from the tagged description $TD$.

@@

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite sets, the
cure lies in rationalising this information. Let us first tackle the
issue of data-type declaration. The grammar of data-types we are used
to can be straightforwardly translated to the ``sigmas-of-sigmas''
paradigm. Hence, a data-type definition can be turned into the
corresponding description. Then, at the level of definitions, we
associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

>>>>

In the simply-typed world, the essence of data-type definition can be
summed up by the expression \emph{sum-of-product}. A data-type is defined
by a finite sum -- a choice -- of constructors, each of them is
composed of a product -- a tuple -- of arguments. This form governs
the grammar of data-type definition in languages such as Haskell or
OCaml.

To model these data-types, we simply need to capture this grammar in
our dependently-typed setting. Here, the notion of sum-of-product
naturally translates into \emph{sigmas-of-sigmas}. However, beyond
this intuition, it should be clear that a direct encoding of
data-types through raw $\Sigma$-types is not a viable option. Indeed,
just as for finite sets, encoding throw away information when we
crucially need it. In the realm of data-types, things become even more
tougher, as one could legitimately expect an induction principle, for
instance. Anonymous in a see of $\Sigma$s, our data-types have little
chance of survival. \note{Ridiculous tentative of poetry. I was
  really tired when I wrote that.}

@@

\[
\begin{array}{ll}
\stk{
\data \Desc : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DUnit          & \Desc \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \Desc} \Desc \\
    \DIndx          & \Desc \To \Desc
\end{array}
}
&
\stk{
\descop{\_\:}{} : \Desc \To \Set \To \Set \\
\begin{array}{@{}ll@{\:=\:\:}ll}
\descop{\DUnit}{& X}        &  \Unit                                       \\
\descop{\DSigma{S}{D}}{& X} &  \SIGMAS{\V{s} : S}{\descop{D~s}{X}}         \\
\descop{\DIndx{D}}{& X}     &  \TIMES{X}{\descop{D}{X}}
\end{array}
}
\end{array}
\]

@@

\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{} ]})}
                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
}\]

@@

\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: X \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons ]})}
                           {[ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]

@@

\[\stk{
\TreeD : \Set \To \Desc \\
\begin{array}{@{}ll}
\TreeD \: X \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
                           { & [ \DUnit \quad (\DSigma{X}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
\end{array}
}\]

@@

Again, this is a clear reminiscence of the sum-of-product style:
we have a finite sum of constructors. Of course, every description can
be expressed in this style, using a singleton as single constructor if
necessary. A tagged description can trivially be turned into a
full-blown description. We note $\toDesc{TD}$ the description computed
from the tagged description $TD$.

@@

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite sets, the
cure lies in rationalising this information. Let us first tackle the
issue of data-type declaration. The grammar of data-types we are used
to can be straightforwardly translated to the sigmas-of-sigmas
paradigm. Hence, a data-type definition can be turned into the
corresponding description. Then, at the level of definitions, we
associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.
